Verification in Isabelle/HOL of a Set-based Algorithm for Computing Strongly Connected Components

### 2022-09-22 cpp/submission.tex
submitted CPP paper

Several different algorithms exist for efficiently computing SCCs of a graph. Perhaps the most famous one is Tarjan's algorithm~\cite{tarjan:depth-first}. It performs a depth-first traversal of the nodes of the graph. When it encounters an edge that leads to an already visited node that is part of the current DFS path, it detects a cycle, and it tracks (via additional information stored for the encountered nodes) which nodes belong to the same SCC. Dijkstra~\cite{dijkstra:finding} suggested a variant of a DFS traversal where the stack contains roots of (partial) SCCs. When a cycle is detected, the corresponding partial SCCs are collapsed, shortening the stack. Both algorithms are efficient because they operate in time $O(|V|+|E|)$ where $V$ and $E$ denote the sets of nodes and edges of the graph, since every node and every edge is visited only once during the traversal, but they differ in their algorithmic structure and in the auxiliary information that must be stored.

However, both these algorithms and their numerous variants are fundamentally sequential because their correctness crucially relies on the graph being traversed in DFS order. Bloemen et al.~\cite{bloemen:strong,bloemen:multi-core} introduced a variant of Dijkstra's algorithm for computing SCCs that can be implemented on a shared-memory multi-core processor, based on efficient data structures for (i) implementing a union-find structure representing partial SCCs and (ii) a cyclic list for iterating over states, that minimize locking. This algorithm has been implemented in the high-performance model checker LTSmin. The correctness proofs in Bloemen's thesis are based on ordinary paper-and-pencil reasoning and therefore do not correspond to the same exacting standards as the machine-checked verification of model checking algorithms mentioned above. The problem is particularly delicate given that a multi-core implementation of the algorithm need not strictly adhere to a DFS traversal order, and any core may revisit nodes that have already been visited by other cores.

The present work makes a first contribution towards a machine-checked proof of the SCC algorithm by Bloemen et al.\ by establishing the correctness of the sequential version of the algorithm. Although this is arguably less challenging than the proof of the parallel version, we consider it to be a necessary first step that establishes a baseline against which the parallel algorithm can be compared. In particular, we exhibit inductive invariants that ensure the correctness of the sequential algorithm, and it then becomes meaningful to consider which invariants can be relaxed in a parallel execution. We are interested in the correctness of the high-level algorithm and therefore do not consider the concrete representation of the data structures on which it relies, or the generation of efficiently executable code implementing the sequential algorithm.

@@

\subsection{Lemmas establishing the pre-conditions}
\label{sec:pre-pre}

@@

\subsection{Pre-conditions imply post-conditions}
\label{sec:pre-post}

@@

@@

\begin{table*}
  \centering
  \begin{tabular}{|l|r|r|}
    \hline
    Category                         &  Definitions   &  Theorems and proofs\\
    \hline\hline
    Precedence on lists              &        2 loc   &         119 loc\\
    \hline
    Graphs and reachability          &       15 loc   &          64 loc\\
    \hline
    Formalization of the algorithm   &       64 loc   &           6 loc\\
    \hline
    Invariants, pre-/post-conditions &       68 loc   &         120 loc\\
    \hline
    Lemmas about \prog{unite}        &                &         663 loc\\
    \hline
    Partial correctness of \prog{dfs} and \prog{dfss} &  &     1575 loc\\
    \hline
    Termination of \prog{dfs} and \prog{dfss} & 13 loc   &      236 loc\\
    \hline
    Total correctness                &                &           5 loc\\
    \hline\hline
    Sum                              &      162 loc   &        2788 loc\\
    \hline
  \end{tabular}
  \caption{Quantitative information about our formalization.}
  \label{tab:effort}
\end{table*}

@@

Table~\ref{tab:effort} summarizes the overall effort of the formalization. We report the number of lines (definitions and theorems/proofs) for each main category, ignoring comments and empty lines. It should be noted that our proofs are written in the relatively verbose Isar language, which favors readability and maintainability compared to shorter, but unreadable tactic applications.

We start by defining the precedence relation between elements of a list, which does not exist in Isabelle's standard library, and proving elementary lemmas about this relation, as described in Section~\ref{sec:isabelle}. In particular, this relation is a partial order if the list does not contain duplicates, which is the case for the stacks in our formalization.

@@

Next, we define the algorithm, including the ``environment'' data structure and the functions \prog{unite}, \prog{dfs}, and \prog{dfss}, cf.\ Section~\ref{sec:formalization}. There is little to prove in this category of the formalization, except for the (trivial) exhaustiveness of the patterns in the recursive definitions, and the transitivity of the relation \prog{sub-env}.

@@

The following two categories represent the main proof effort of this work. We start by proving auxiliary lemmas about the function \prog{unite}, and in particular the fact that it preserves the main invariant. The partial correctness proofs establish the pre-conditions for each recursive function call, as well as the post-conditions of the functions, assuming their pre-conditions, as reported in Sections~\ref{sec:pre-pre} and~\ref{sec:pre-post}.

@@

As already noted in the introduction, we are not the first who prove the correctness of an algorithm for computing the SCCs of a graph using a proof assistant. Peter Lammich~\cite{lammich:gabow} used Isabelle/HOL for proving the correctness of Gabow's SCC algorithm. Like Bloemen's algorithm, Gabow's algorithm applies depth-first search and augments partial SCCs whenever a back loop to an already visited vertex of the graph is found. Lammich's proof focuses on efficient code generation, based on the use of his refinement framework~\cite{lammich:refinement} formalized in Isabelle. In this framework, algorithms are represented as monadic functions, and concrete data structures are introduced, for example for representing partial SCCs. In a separate work, Lammich and Neumann contributed significantly to the development of a fully verified executable model checker~\cite{esparza:cava} by developing a framework for verifying depth-first search algorithms~\cite{lammich:depth-first}, including Tarjan's algorithm. In contrast to these works, our proof is based on a more abstract representation of Bloemen's algorithm and its data structures, which leads to higher-level invariants. Indeed, our main objective is not to obtain an efficient sequential implementation of Bloemen's algorithm but rather to pave the way for a verification of the parallel version of that algorithm.

Chen et al.~\cite{chen:tarjan} present a comparison of the proofs of Tarjan's algorithm for computing the SCCs of a graph in the proof assistants Coq, Isabelle, and Why3. The algorithm is represented at a similar level of abstraction as in our proof, leading to comparable proof effort. However, although Tarjan's algorithm is also based on depth-first search, its algorithmic structure is quite different because no partial SCCs are computed along the way, and a maximal SCC is popped from the stack only when a node is determined to be an SCC root at the end of the function call. The primary focus of the presentation in~\cite{chen:tarjan} is on a comparison between the styles of proof and the overall effort required in the three proof assistants.

@@

In this paper, we reported on the successful proof in Isabelle/HOL of an efficient sequential SCC algorithm proposed by Bloemen et al.~\cite{bloemen:strong}. In particular, we exhibited the inductive invariant \prog{wf-env} that is preserved throughout the execution of the algorithm, as well as pre- and post-conditions for the two mutually recursive functions representing the algorithm. Although our proof is based on arguments given in the correctness proof in Bloemen's thesis, we had to add non-trivial predicates in order to obtain an inductive invariant. For example, the paper-and-pencil proof does not mention the facts that the partial SCCs maintained by the algorithm are maximal w.r.t.\ the current state of knowledge, or that the root of an SCC is the oldest node in depth-first order, but these are necessary for the argument to be conclusive.

As we mentioned previously, the main objective of our work is to serve as a first step towards a machine-checked proof of a parallel implementation of Bloemen's algorithm on a multi-core processor. We therefore formulated our correctness predicates at a high level of abstraction, without considering how the data manipulated by the algorithm would actually be represented in an implementation. We hope that in this way, it will be possible to consider which relaxations of our invariants are possible in a parallel execution of the algorithm. Preliminary experiments with encoding the parallel version in TLA$^+$ and checking several plausible invariants of the algorithm over small graphs using model checking indicate that some invariants break without corrupting the overall result of the SCC computation~\cite{vandepol:exploring}, but that slight modifications of the algorithm are incorrect. We believe that the invariants that we identified for the sequential algorithm, together with the detailed proof, will provide a sound basis for extending the proof to the parallel version.

% As a result of this whole year of work, we successfully proved correctness of an efficient sequential set-based SCC algorithm. In comparison to Tarjan's algorithm, a formal proof of which was written by Chen et. al \cite{chen:tarjan}, this algorithm collapses loops as soon as they are found, thus minimizing the stack size. Our formalization and  proof in Isabelle are quite similar to theirs as they are based on higher-level representations of the algorithms as elementary recursive functions. Hence, our invariants are quite simple compared to those expressed in Lammich's work \cite{lammich:gabow} since he exploits the refinement framework for deriving executable code, resulting in implementable-like representations that make expressing invariants more difficult from a higher-level point of view.

% \begin{itemize}
% \item successfully proved correctness of an efficient sequential SCC algorithm
% \item compare to Tarjan proof (Chen et al.): both are DFS algorithms, the algorithm considered here collapses loops as soon as they are found, minimizing stack size
% \item compare to Lammich's work: he exploits the refinement framework for deriving executable code, we work on a higher-level representation of the algorithm as elementary recursive functions, hence invariants are simpler
% \item objective: stepping stone towards a proof of the parallel version of the algorithm from Bloemen's thesis -- understand which invariants can be relaxed.
% \end{itemize}

>>>>

Several different algorithms exist for efficiently computing SCCs of a graph. Perhaps the most famous one is Tarjan's algorithm~\cite{tarjan:depth-first}. It performs a depth-first traversal of the nodes of the graph. When it encounters an edge that leads to an already visited node that is part of the current DFS path, it detects a cycle, and it tracks
% (via additional information stored for the encountered nodes)
which nodes belong to the same SCC. Dijkstra~\cite{dijkstra:finding} suggested an algorithm that is also based on a DFS traversal of the graph. In this algorithm, the stack contains roots of (partial) SCCs. When a cycle is detected, the corresponding partial SCCs are collapsed, shortening the stack. Both algorithms are efficient because they operate in time $O(|V|+|E|)$ where $V$ and $E$ denote the sets of nodes and edges of the graph, since every node and every edge is visited only once during the traversal, but they require different auxiliary information to be stored during execution.

Crucially, both these algorithms and their numerous variants are fundamentally sequential because their correctness relies on the graph being traversed in DFS order. Bloemen et al.~\cite{bloemen:strong,bloemen:multi-core} introduced a variant of Dijkstra's algorithm for computing SCCs that can be implemented on a shared-memory multi-core processor, and they designed efficient data structures for (i) implementing a union-find structure representing partial SCCs and (ii) a cyclic list for iterating over states, that minimize locking. Their algorithm has been implemented in the high-performance model checker LTSmin. They empirically demonstrate that the algorithm ensures good speed-up even for graphs that have few large SCCs, as is often the case in model checking. The correctness proofs in Bloemen's thesis are based on ordinary paper-and-pencil reasoning and therefore do not correspond to the same exacting standards as the machine-checked verification of model checking algorithms mentioned above. The problem is particularly delicate given that a multi-core implementation of the algorithm need not strictly adhere to a DFS traversal order, and any core may revisit nodes that have already been visited by other cores, which might not only be an efficiency problem but also jeopardize correctness.

The present work makes a first contribution towards a machine-checked proof of Bloemen et al.'s SCC algorithm by establishing the correctness of the sequential version of the algorithm. Although this is arguably less challenging than the proof of the parallel version, we consider it to be a necessary first step that establishes a baseline against which the parallel algorithm can be compared. In particular, we exhibit inductive invariants that ensure the correctness of the sequential algorithm, and it then becomes meaningful to consider which invariants can be relaxed in a parallel execution. We are interested in the correctness of the high-level algorithm and therefore do not consider the concrete representation of the data structures on which it relies, or the generation of efficiently executable code implementing the sequential algorithm.

@@

\subsection{Lemmas establishing the pre-conditions}
\label{sec:pre-implies-pre}

@@

\subsection{Pre-conditions imply post-conditions}
\label{sec:pre-implies-post}

@@

\begin{table*}
  \centering
  \begin{tabular}{|l|r|r|}
    \hline
    Category                         &  Definitions   &  Theorems and proofs\\
    \hline\hline
    Precedence on lists              &        2 loc   &         119 loc\\
    \hline
    Graphs and reachability          &       15 loc   &          64 loc\\
    \hline
    Formalization of the algorithm   &       64 loc   &           6 loc\\
    \hline
    Invariants, pre-/post-conditions &       68 loc   &         120 loc\\
    \hline
    Lemmas about \prog{unite}        &          ---   &         663 loc\\
    \hline
    Partial correctness of \prog{dfs} and \prog{dfss} & --- &  1575 loc\\
    \hline
    Termination of \prog{dfs} and \prog{dfss} & 13 loc   &      236 loc\\
    \hline
    Total correctness                &          ---   &           5 loc\\
    \hline\hline
    Sum                              &      162 loc   &        2788 loc\\
    \hline
  \end{tabular}
  \caption{Quantitative information about our formalization.}
  \label{tab:effort}
\end{table*}

@@

@@

Table~\ref{tab:effort} summarizes the overall effort of the formalization. We report the number of lines (definitions and theorems/proofs) for each main category, ignoring comments and empty lines. It should be noted that our proofs are written in the relatively verbose Isar language, which favors readability and maintainability compared to shorter, low-level tactic applications.

We start by defining the precedence relation between elements of a list, which is not pre-defined in Isabelle's standard library, and proving elementary lemmas about this relation, as described in Section~\ref{sec:isabelle}. In particular, this relation is a partial order if the list does not contain duplicates, which is the case for the stacks in our formalization.

@@

Next, we define the algorithm, including the ``environment'' data structure and the functions \prog{unite}, \prog{dfs}, and \prog{dfss}, cf.\ Section~\ref{sec:formalization}. This category of our formalization mainly consists of definitions and few proofs, except for the (trivial) exhaustiveness of the patterns in the recursive definitions, and the transitivity of the relation \prog{sub-env}.

@@

The following two categories represent the main proof effort of this work. We start by proving auxiliary lemmas about the function \prog{unite}, and in particular the fact that it preserves the main invariant. The partial correctness proofs establish the pre-conditions for each recursive function call, as well as the post-conditions of these calls, assuming their pre-conditions, as reported in Sections~\ref{sec:pre-implies-pre} and~\ref{sec:pre-implies-post}.

@@

As we already noted in the introduction, we are not the first who prove the correctness of an algorithm for computing the SCCs of a graph using a proof assistant. Peter Lammich~\cite{lammich:gabow} used Isabelle/HOL for proving the correctness of Gabow's SCC algorithm. Like Bloemen's algorithm, Gabow's algorithm is based on Dijkstra's algorithm: it applies depth-first search and combines partial SCCs whenever a back loop to an already visited vertex of the graph is found. Lammich's development focuses on efficient code generation, based on the use of his refinement framework~\cite{lammich:refinement} formalized in Isabelle. In this framework, algorithms are represented as monadic functions, and concrete data structures are introduced, for example for representing partial SCCs. In separate work, Lammich and Neumann contributed significantly to the development of a fully verified executable model checker~\cite{esparza:cava} by developing a framework for verifying depth-first search algorithms~\cite{lammich:depth-first}, including Tarjan's algorithm. In contrast to these works, our proof is based on a more abstract representation of Bloemen's algorithm and its data structures, which leads to higher-level invariants. Indeed, our main objective is not to obtain an efficient sequential implementation of Bloemen's algorithm but rather to pave the way for a verification of the parallel version of that algorithm.

Chen et al.~\cite{chen:tarjan} present a comparison of the proofs of Tarjan's algorithm for computing the SCCs of a graph in the proof assistants Coq, Isabelle, and Why3. The algorithm is represented at a similar level of abstraction as in our proof, leading to comparable proof effort. However, although Tarjan's algorithm is also based on depth-first search, its algorithmic structure is quite different because no partial SCCs are computed along the way, and a maximal SCC is popped from the stack only when a node is determined to be an SCC root at the end of the DFS search starting at that node. The primary focus of the presentation in~\cite{chen:tarjan} is on a comparison between the styles of proof and the overall effort required in the three proof assistants.

@@

In this paper, we reported on the successful proof in Isabelle/HOL of an efficient sequential SCC algorithm proposed by Bloemen et al.~\cite{bloemen:strong,bloemen:multi-core}. In particular, we exhibited the inductive invariant \prog{wf-env} that is preserved throughout the execution of the algorithm, as well as pre- and post-conditions for the two mutually recursive functions representing the algorithm. Although our proof is based on arguments given in Bloemen's thesis, we had to add non-trivial predicates in order to obtain an inductive invariant. For example, the paper-and-pencil proof does not mention the facts that the partial SCCs maintained by the algorithm are maximal w.r.t.\ the current state of knowledge, or that the root of an SCC is the oldest node in depth-first order, but these are necessary for the argument to be conclusive.

As we mentioned previously, the main objective of our work is to serve as a first step towards a machine-checked proof of a parallel version of Bloemen's algorithm that can be implemented on a multi-core processor. We therefore formulated our correctness predicates at a high level of abstraction, without considering how the data manipulated by the algorithm would actually be represented in a program. We hope that this will help considering which relaxations of our invariants are possible in a parallel execution of the algorithm. Preliminary experiments with encoding the parallel version in TLA$^+$ and checking several plausible invariants of the algorithm over small graphs using model checking indicate that some invariants break, but that the overall result of the SCC computation is not corrupted~\cite{vandepol:exploring}. However, apparently minor modifications of the algorithm were found to be incorrect. We believe that the invariants that we identified for the sequential algorithm, together with the detailed proof, will provide a sound basis for extending the proof to the parallel version.
